---
title: "Running Back in Style"
format: html
---

Running Backs had their moment this past season. A recently maligned position group thought to provide little value to team production made a momentary sigh of relief. Leading the comeback story was Saquon Barkley's near record breaking campaign and Super Bowl run after being discarded by his former team and now division rival New York Giants. His monumental impact to the Eagles Super Bowl season did not go unnoticed as he was rewarded with a big [contract extention](https://www.espn.com/nfl/story/_/id/44124382/nfl-free-agency-2025-saquon-barkley-change-running-back-contracts-eagles) in March. He wasn't the only one rewarded with a new deal as Derrick Henry received a 2 year extentions following his recent debut season with the Ravens. Those players however aren't anomalies, as some exciting and dynamic stars such as Detroit's Jamyhr Gibbs and Atlanta's Bijan Robinson have both added to the strong case of the position group's value revival. It seems to have also convinced teams this season to invest more in the short dip in the RB market. 

In this year's draft, 5 RBs were selected in the top 60 picks, 1 of which (Ashton Jeanty) was selected number 6. All of this leads me to believe that RB value is on the upswing. 


For the past few years, the analytics has discussed at length, the value of each position driven by a desire to optimize where teams spend their resources (draft picks, money, etc.). The prevailing argument coming out of this research was that teams can be successful without investing heavily in position due to a variety of factors including teams adopting pass-heavy approaches, importance of offensive line play and impact, backfield by committee, and the contract values of running backs.

I'm hoping to address the last 2 points in this post. The decline of a "workhorse" back and the increase in rush duties shared among multiple runnings and, at times, the QB. Teams have tended to rely slightly less on one RB for a number of reasons. One of the prime reasons was the hesitancy to put the that kind of load on the bodies of one player. Football is a violent game and taking hits has a toll. A recent run of talented RBs declining after signing massive contracts is a reason teams have been scared off of committing large amounts of resources recently. Players like Todd Gurley, Ezekiel Elliot, and Le'Veon Bell have all experienced precipotous decline and/or serious injuries after signing large long-term contracts.

While it's certainly proven riskier to spend resources on runnings there still have been plenty of RBs who have continued to produce. Henry himself had a stellar season at age 30. Alvin Kamara, Christian, McCafferey, Josh Jacobs, and Barkley have all produced well after receiving long-term contracts. How can we predict whether the contract is worth it? Is there a way to tell how much proverbial gas is left in the tank for these guys. Let's find out...


```{r}
# Create Time Weighting 

weighted_sum = function(x,alpha=0.8){
  
  # Take in sliding window of vector of chosen metric

  n <- length(x)
  w_sum = sapply(1:n, function(i) {
    weights <- alpha^((i - 1):0)  # Most recent gets highest weight (alpha^0 = 1)
    values <- x[1:i]
    weighted_sum <- sum(values * weights, na.rm = FALSE) / sum(weights)
    return(weighted_sum)
  })
  
  return(w_sum)
  
}
```


We'll load our relevnat data in this section. Hypothesizing which types of stats could be the biggest influencers of running back carries, we'll break things down into previous success, injuries, and monetary commitment.

```{r}
library(nflreadr)
library(tidyverse)

player_stats = load_player_stats(seasons = seq(2006,most_recent_season()),stat_type = "offense")
player_rosters = load_rosters(seasons = seq(2006,most_recent_season()))


# Injuries

nflreadr::load_injuries(seasons = seq(2009,most_recent_season())) -> injuries

injuries |> 
  filter(position=="RB") |> 
  group_by(gsis_id,season) |> 
  summarise(weeks_missed = sum(report_status == "Out",na.rm = T)) |> 
  ungroup() -> rb_injuries

# Aggregate and Join
stat_cols = c('carries','rushing_yards','targets','receptions','receiving_yards',"fantasy_points","games")
player_stats |> 
  filter(position == 'RB') |>
  arrange(desc(season),desc(week)) |>
  group_by(player_id,season,) |> 
  summarise(games = n(),
            # Team they start season with
            recent_team = first(recent_team),
            player_name = last(player_name),
            across(stat_cols,.fns = sum)) |> 
  select(c(player_id,player_name,recent_team,season,stat_cols)) -> rush_stats


# Calculate cumulative sums of stats endured so far
rush_stats |> 
  group_by(player_id) |> 
  left_join(rb_injuries,by = c('season','player_id' = 'gsis_id')) |> 
  replace_na(list(weeks_missed = 0)) |> 
  # Compute 1 year lag of all numeric columns
  mutate(across(where(is.numeric), ~lag(.), .names = "{.col}_lag1"),
         # Cumulative sum of injuries
         weeks_inj = lag(cumsum(weeks_missed)),
         across(stat_cols, ~lag(cumsum(.)), .names = "{.col}_career_sum"),
         across(stat_cols, ~lag(weighted_sum(.)), .names = "{.col}_wt_sum")) -> agg_stats
  
```


#### Depth Charts 

```{r}
player_rosters |> 
  filter(position == 'RB') |> 
  arrange(season,week) |> 
  group_by(gsis_id) |> 
  # Some NAs for draft number
  mutate(draft_number = min(draft_number,na.rm = T)) |> 
  group_by(season,gsis_id,position,birth_date,entry_year,college,draft_club,draft_number) |> 
  summarise(full_name = last(full_name),
            week = last(week),
            team = last(team),
            years_exp = last(years_exp),
            height = last(height),
            weight = last(weight),
            avg_depth = mean(depth_chart_position)) |> 
  ungroup() |> 
  mutate(
    age = round(lubridate::time_length(as.Date(paste(season,'03','01',sep = '-'))-birth_date,unit = 'years')),
    # Handles UDFA
    draft_number = as.numeric(ifelse(
      is.na(draft_number),263,draft_number
      
    ))) |> 
  arrange(desc(season)) |> 
  select(gsis_id,season,years_exp,height,weight,age,draft_number)-> rb_roster_info

rb_roster_info |> 
  filter(gsis_id == "00-0032780")
```


#### Contract Info

```{r}
# Contract Info
nflreadr::load_contracts()-> contracts

contracts |> 
  # Last 
  filter(position=='RB') |> 
  select(year_signed,gsis_id,apy,apy_cap_pct) |> 
  # Grab max contract value per player per year (Players may sign multiple contracts per year)
  group_by(gsis_id,year_signed) |> 
  mutate(apy = max(apy),apy_cap_pct = max(apy_cap_pct)) |> 
  distinct() -> rb_contracts

# Check no one has more than 1 contract per year
rb_contracts |> 
  group_by(year_signed,gsis_id) |> 
  tally() |> 
  arrange(desc(n))

roster_contract = rb_roster_info |> 
  left_join(rb_contracts,by = c("season" = "year_signed","gsis_id")) |> 
  arrange(desc(season)) |> 
  group_by(gsis_id) |> 
  # Fill contract columns "up" to fill in blanks
  fill(c(apy_cap_pct),.direction = 'up')


```


Let's take a look at the salary cap space devoted to RBs by age.

```{r}

roster_contract |> 
  group_by(age) |> 
  summarise(avg_apy = mean(apy_cap_pct,na.rm = T)) |> 
  ggplot(aes(x = age,y = avg_apy)) +
  geom_col() +
  labs(title = "Peak Cap Space Hit")

```
Ok, looks like running backs typically take up the largest % of the cap by their age 27 season. This isn't a foolproof plan, as these estimates can be sensitive to small samples, but we're trying to capture year to year changes.

```{r}
agg_stats |> 
  left_join(y = roster_contract,by = c("season","player_id" = "gsis_id")) |> 
  filter(season >= 2009) |> 
  # Create a new target: "Total Touches"
  mutate(touches = carries + receptions) |> 
  arrange(desc(season)) |> 
  # Select relevant columns for modeling
  select(1:4,ends_with("_sum"),ends_with("_lag1"),"touches",36:44,-c(apy))-> model_data

# Select base

```


### Data Splitting

Before we start with any analysis, we should start with our preprocessing steps so we're not leaking any information to a model.

```{r}
library(tidymodels)
set.seed(123)
# Split the data
splits <- rsample::initial_split(model_data, prop = 0.8)
rb_train <- rsample::training(splits)
rb_test <- rsample::testing(splits)
```


Create recipes (eventually we will compare).

```{r}
# Create cross-validation folds
rb_folds <- vfold_cv(rb_train, v = 5)

some_vars <- names(model_data)[5:ncol(model_data)]

# Define preprocessing recipes
# Recipe 1: Basic preprocessing
basic_recipe <- recipe(touches ~ ., data = rb_train) |> 
  update_role(season, new_role = "ID") |> 
  update_role(player_id, new_role = "ID") |> 
  step_select(all_of(some_vars)) |> 
  step_normalize(all_numeric_predictors()) |> 
  step_corr(all_numeric_predictors(), threshold = 0.9) |> 
  step_impute_median(apy_cap_pct) |> 
  step_naomit(all_numeric_predictors()) |> 
  step_filter_missing(
    threshold = 0.3
  ) 
  

pca_recipe <- basic_recipe |> 
    step_pca(all_numeric_predictors(), threshold = 0.95)


# Define model specifications
# Linear regression
lm_spec <- linear_reg() |>
  set_engine("lm")
  

# Random Forest
rf_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
) |>
  set_engine("ranger") |>
  set_mode("regression")

# Elastic Net
glm_spec <- linear_reg(
  penalty = tune(),
  mixture = tune()
) |>
  set_engine("glmnet")

# XGBoost
xgb_spec <- boost_tree(
  mtry = tune(), trees = 1000, min_n = tune(), tree_depth = tune(),
  learn_rate = tune(), loss_reduction = tune(), sample_size = NULL,
  stop_iter = tune()
) |>
  set_engine("xgboost") |>
  set_mode("regression")

prep(basic_recipe)

```


```{r}
# Create workflow set
# Combine different recipes with different models
model_set <- workflow_set(
  preproc = list(
    basic = basic_recipe,
    pca = pca_recipe
  ),
  models = list(
    lm = lm_spec,
    rf = rf_spec,
    glm = glm_spec,
    xgb = xgb_spec
    ),
  cross = TRUE  # Create all combinations
)

# View the workflow set
print(model_set)
```


```{r}
#doParallel::registerDoParallel()

model_results <- model_set |>
  workflow_map(
    fn = "tune_grid",
    resamples = rb_folds,
    grid = 10,  # Number of hyperparameter combinations to try
    metrics = metric_set(rmse, rsq, mae),
    verbose = TRUE,
    seed = 123
  )

# View results summary
# Collect and rank metrics
model_rankings <- model_results |>
  rank_results(rank_metric = "rmse", select_best = TRUE) |>
  filter(.metric == "rmse")

print(model_rankings)
```

Okay, it looks like the random forrest regressor was the best at predicting new samples. It also appears that our PCA did not do a great job of minimizing enough of the information in the feature space.

```{r}
# Visualize model performance
autoplot(model_results, metric = "rmse") +
  theme_minimal() +
  labs(title = "Model Performance Comparison",
       subtitle = "Lower RMSE indicates better performance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

best_workflow_id <- model_rankings |> 
  slice_head(n = 1) |> 
  pull(wflow_id)

model_rankings
```

This autoplot gives us some added context to our tested model performance. While it looks like the tree-based models give us slightly better fit than the glms. It can be argued that the performance improvement of the random forest model doesn't justify the increased complexity and training time required. However, implementing that RF forest shouldn't be too difficult in practice. We'll stick with that.


```{r}
best_workflow <- model_results |> 
  extract_workflow(best_workflow_id)

# Finalize the best model with optimal hyperparameters  
best_results <- model_results |> 
  extract_workflow_set_result(best_workflow_id)
 
best_params <- best_results |> 
  select_best(metric = "rmse")

final_workflow <- best_workflow |> 
  finalize_workflow(best_params)

# Fit the final model on training data and evaluate on test data
final_fit <- final_workflow |> 
  last_fit(splits)

final_workflow
```


Let's see how well our model approximates the number of "touches" a ball carrier is expected to take.

```{r}
# Test set performance
test_performance <- final_fit |>
  collect_metrics()

print("Test Set Performance:")
print(test_performance)

# Predictions vs actual plot
final_fit |>
  collect_predictions() |>
  ggplot(aes(x = touches, y = .pred)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Actual Touches", y = "Predicted Touches",
       title = "Best Model: Predictions vs Actual") +
  theme_minimal()
```


```{r}
# Create a detailed comparison plot
model_results |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(aes(x = reorder(wflow_id, mean), y = mean)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), 
                width = 0.1) +
  coord_flip() +
  labs(x = "Workflow", y = "RMSE",
       title = "Model Performance Comparison with Error Bars") +
  theme_minimal()

```
While we have our model prepared, we're still missing something we didn't get a chance to address in the last few posts. We typically throw out the first week/year/date (not recommended in practice) since we had no data. Well we still don't, but that want stop us from making educated guesses. 


### Imputation

As you saw, I took the lag value of several variables meaning that we cannot make predictions for running backs with 0 years of experience, aka rookies. However, I for one would love to see how many carries Ashton Jeanty may reasonably take this year and it should be simple enough to at least TRY to estimate that.

```{r}
rb_train |>
  filter(years_exp == 0) -> rookie_train

rb_train |> 
  filter(years_exp == 0)  -> rookie_test
rookie_train |> 
  ggplot(aes(x = draft_number, y= touches)) +
  geom_point() + 
  geom_smooth(method = "gam")
```


```{r}
# Recipe 1: Rookies
rookie_recipe <- recipe(touches ~ ., data = rookie_train) |> 
  step_normalize(all_numeric_predictors()) |>
  step_corr(all_numeric_predictors(), threshold = 0.9)
```



Okay, not exactly linear by any means, but in-line with our intuition that lower draft picks will get more touches on the ball. You might have noticed that we're using "gam" as the smoothing method. This is to implement a Generalized Additive Model. We won't get into the nitty gritty of what that is in this post, but it's important to know that we use it when we know a relationship between predictor and response in NOT linear.

Below, we'll actually create a GAM model to serve as a our model for rookies.We're going to do the quick and dirty way of building a model without any cross-validation or much preprocessing since that is handled for us.


```{r}
library(mgcv)

# Fit our baseline model - simple linear regression
rookie_fit <- lm(touches ~ season + draft_number + height*weight, data = rookie_train)

# Fit a GAM with 1 predictor x and response y - This was tinkered with
gam_model <- gam(touches ~ s(draft_number) + s(weight) + height + season, data = rookie_train)

# Some more tinkering just cause
gam_model2<- gam(touches ~ s(draft_number) + te(weight, height) + weight,data = rookie_train)

broom::glance(rookie_fit)

broom::glance(gam_model)

summary(gam_model)

summary(gam_model2)

tidy(gam_model, conf.int = TRUE)

# Make predictions on test set
augment(gam_model,rookie_test) |> 
  select(player_id,player_name,recent_team, season,touches,.fitted,.resid,draft_number,height,weight) -> res_rookie

res_rookie |> 
  arrange(desc(season),desc(.resid))

# SSE
sqrt(mean((res_rookie$.fitted-res_rookie$touches)**2))
rmse(res_rookie, touches,.fitted)


res_rookie |> 
  ggplot(aes(x = draft_number, y = abs(.resid))) +
  geom_point()+
  geom_smooth(method = "lm")
```
When we say the model was "tinkered with" it just means we were checking significance with different combinations of variables.


Now that we've finished the GAM model we can combine both datatsets and check what the results tell us.

One of the core features of the GAM application, is the allowance of both linear and non-linear terms. In the output above, we can assume a linear relationship betweens between height and touches and season and touches.


```{r}
plot(ggeffects::ggpredict(gam_model))
gratia::draw(gam_model)
```




```{r}
# Create cross-validation folds
rb_folds <- vfold_cv(rb_train, v = 5, strata = age)

# Define preprocessing recipes
# Recipe 1: Basic preprocessing
basic_recipe <- recipe(carries ~ ., data = rb_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_corr(all_numeric_predictors(), threshold = 0.9)

# Recipe 2: More complex preprocessing with polynomial terms
poly_recipe <- recipe(mpg ~ ., data = rb_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_poly(hp, wt, degree = 2) |>
  step_corr(all_numeric_predictors(), threshold = 0.9)

# Recipe 3: PCA preprocessing
pca_recipe <- recipe(mpg ~ ., data = rb_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), threshold = 0.95)

# Define model specifications
# Linear regression
lm_spec <- linear_reg() |>
  set_engine("lm")

# Random Forest
rf_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
) |>
  set_engine("ranger") |>
  set_mode("regression")

```




### Missingness

```{r}
rb_train |> 
  ungroup() |> 
  miss_var_summary()
```


Seems as though there are a decent number of missing values for that apy contract value. 

```{r}
library(naniar)
library(tidyr)

model_data |>
  nabular() |>
  ggplot(aes(x = season,
             y = carries,
             fill = apy_cap_pct_NA)) + 
  geom_col()

# Draft Number
model_data |>
  nabular() |>
  ggplot(aes(x = draft_number,
             y = apy_cap_pct)) + 
  geom_miss_point() +
  facet_wrap(~season)


```


```{r}
# impute
gam(apy_cap_pct ~ s(fantasy_points_wt_sum) + s(draft_number) + s(years_exp) +
     te(games_career_sum,age)+
   +s(season_lag1)+
     s(fantasy_points_wt_sum)+
     s(weight)+
     s(weeks_inj), 
   data = rb_train,
   method="REML",
   select=TRUE)-> impute_cap_pct

summary(impute_cap_pct)
```

### Aging Curves


Using carries as a proxy for success. We run into the following problem as explained by the graph below.

```{r}
model_data |> 
  group_by(age = round(approx_age)) |> 
  summarise(avg_carries = mean(carries,na.rm = T)) |> 
  ungroup() |> 
  ggplot(aes(x = age,y=avg_carries))+
  geom_smooth()
```
Interesting, is it reasonable to expect an upward trajectory the entire way? Perhaps early on in their careers, but I hardly expect them to reach their peak at age 37! This is probably due to a cognitive bias called *survivorship bias*. It's when our sample mistakenly represents only the "success" of population rather than the entire population.



Links:
https://opensourcefootball.com/posts/2020-08-25-open-source-fantasy-football-visualizing-trap-backs/

https://www.unexpectedpoints.com/p/what-analytical-draft-value-curves

https://opensourcefootball.com/posts/2023-02-23-nfl-draft-value-chart/