---
title: "Predicting Wide Receiver Fantasy Points w/ Tidymodels Pt 2"
subtitle: "Mixed Models"
author: "Sameer Sapre"
date: "2024-09-16"
categories: [code, analysis]
#draft: TRUE
editor: visual
---

I made a slight mistake. I broke a core assumption of linear regression that I made in the previous edition of the model. To use linear regression, one must assume that each data point is **independent** of one another. However, this is not the case in the dataset I was working with. Take a look at the sample below. Here are some of the league leaders in receptions for 2023. We can see clearly that we are dealing with some repeated measurements.

```{r}
ma_wr %>%
  filter(season == 2023) %>%
  group_by(player_id) %>%
  mutate(season_catches = sum(receptions)) %>%# %>%
  ungroup() %>%
  arrange(desc(season_catches)) %>%
  top_n(170) %>%
  ggplot(aes(x = week, y = fantasy_points)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~ player_name) 
    

```

We can see here, that there are several records of Aiyuk in 2023 (even more if we include other seasons). This is an example of what scientists may call "repeated" measures. It's a phrase to describe the process of taking several measurements of a variable on the same subject (Aiyuk). This happens for almost every receiver in our dataset.

```{r}
exp_recipe = train_data %>%
  recipe(formula = fantasy_points ~ .) %>%
  update_role(c(recent_team,log_fantasy_points,fantasy_points_ppr,fantasy_points_target),new_role = 'ID') %>%
  # Generally not recommended to throw out all data, but for brevity, let's remove NAs
  step_impute_median(all_numeric_predictors()) %>%
 # Remove zero variance predictors (ie. variables that contribute nothing to prediction)
  step_zv(all_predictors()) %>%
  step_center(all_numeric_predictors()) 


```

```{r}
# Calculate the correlation matrix wrt to log fantasy points
cor_matrix <- train_data %>% 
    select(starts_with("wt_"), fantasy_points) %>%
    cor(use = "complete.obs")

cor_data <- reshape2::melt(cor_matrix)

# Sort variables by correlation value
cor_data %>% filter(Var1 == "fantasy_points") %>% select(-Var1) %>% arrange(desc(value))

```

```{r}
library(lme4)
library(multilevelmod)


mixed_mod_spec <- linear_reg() %>%
  set_engine("lmer")

# Setup workflow
glm_wflow <- workflow() %>% 
  add_recipe(exp_recipe) #%>%
  #add_model(spec = mixed_mod_spec, formula = fantasy_points ~ 1 + (1+ wt_wopr|player_id) + wt_wopr)

# Intercept Only
glm_wflow %>%
  add_model(spec = mixed_mod_spec, formula = fantasy_points ~ 1 + (1|player_id)) %>%
  fit(data = train_data) %>%
  extract_fit_engine() -> int_only

print('Intercept ONly')
summary(int_only)

# Mixed Effects
glm_wflow %>%
  add_model(spec = mixed_mod_spec, formula = fantasy_points ~ 1 + (1|player_id) + wt_wopr + wt_depth_team) %>%
  fit(data = train_data) %>%
  extract_fit_engine() -> mixed_mod

print('Mixed Effect')
summary(mixed_mod)

# Fit 2
glm_wflow %>%
  add_model(spec = mixed_mod_spec, formula = fantasy_points ~ 1 + (1+wt_wopr|player_id) + wt_wopr) %>%
  fit(data = train_data) %>%
  extract_fit_engine()-> ranef_mod
  
print("Random Effect")
summary(ranef_mod)
  
# Create cross validation splits
#wr_folds <- vfold_cv(train_data, v = 5)
```
